\documentclass[10pt,fleqn]{article}
\usepackage{hyperref}


\setlength{\topmargin}{-.75in}
\addtolength{\textheight}{2.00in}
\setlength{\oddsidemargin}{.00in}
\addtolength{\textwidth}{.75in}

\nofiles

\pagestyle{empty}

\setlength{\parindent}{0in}

\input{/cygdrive/m/tex/commands/commands}

\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vskip0.1in\hrule\vskip0.1in
\noindent
{\bf Math 4610 Fundamentals of Computational Mathematics  - Lecture 2.} 
\vskip0.1in\hrule\vskip0.1in
\noindent
In this lesson, you will be introduced to simple algorithms for determining
\lq\lq machine epsilon\rq\rq\ to quantify computer number precision. The output
of these algorithms serve as a warning about using computers in the solution of
real mathematical problems. In adition to finite precision in the representation
of numbers on a computer, most approximation algorithms will require large
numbers arithmetic operations to determine an appropriate approximation.

Representation of round off error and the analysis of how errors accumulate in
complex computations. One way to proceed is through interval analysis applied to
numbers that are approximated in machine arithmetic. A simple 
example of how to build a shared library will also be discussed. This is an
important skill to have in computational mathematics.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vskip0.1in\hrule\vskip0.1in
\noindent
{\bf Content Items:}
\begin{list}{$\bullet$}{\usecounter{beans} \parsep=0pt \listparindent=0pt
\topsep=0pt \rightmargin=.35in \leftmargin=.35in \labelsep=5 pt
\itemsep=2pt}
  \item {\bf Building a Repository on Github:} - We will discuss how to create a
     general repository to start class. The idea is that students should create
     the math4610 repository in this part of the lecture.
     \href{https://jvkoebbe.github.io/math4610/lectures/lecture_02/md/newrepository_primer}{$>$ $>$ go there}
     \href{https://jvkoebbe.github.io/math4610/lectures/lecture_02/pdf/newrepository_primer.pdf}{(pdf)}
  \item {\bf Homework repositories on Github:} - We will discuss how to create a
     homework repository that each student must have to complete homework. The
     homework repository should be a private repository with you instructor as
     the only collaborator. Homework will be graded by first cloning the
     repository and then your constructor can create a branch for grading. This
     can be repeated by creating a pull request on the repository.
     \href{https://jvkoebbe.github.io/math4610/lectures/lecture_02/md/homework_description}{$>$ $>$ go there}
     \href{https://jvkoebbe.github.io/math4610/lectures/lecture_02/pdf/homework_description.pdf}{(pdf)}
  \item {\bf Practical Intuition about Machine Precision:} Some simple examples
     of how computers represent numbers are given that indicate that almost all
     of the work done by computers is an approximation. Finite representation in
     a binary number system has significant limitations. Since this is what we
     have to use, it is important to understand the implications of finite
     number representation.
     \href{https://jvkoebbe.github.io/math4610/lectures/lecture_02/html/error_practical_intuition.html}{$>$ $>$ go there}
     \href{https://jvkoebbe.github.io/math4610/lectures/lecture_02/pdf/error_practical_intuition.pdf}{(pdf)}
  \item {\bf Code for Determining Machine Precision:} A simple algorithm for
     determining the precision for number representation will be covered. Two
     different versions of the code will be presented that determine single and
     double precision for your computer. Note that these two codes will be used
     to show students how to build a shared library.
     \href{https://jvkoebbe.github.io/math4610/lectures/lecture_02/md/maceps}{$>$ $>$ go there}
     \href{https://jvkoebbe.github.io/math4610/lectures/lecture_02/pdf/maceps.pdf}{(pdf)}
  \item {\bf Wrap up and Questions:} If there is time and anyone has questions
        about the lecture, these will be addressed.
\end{list}
\vskip0.1in\hrule\vskip0.1in
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}










  \item {\bf Using Version Control Systems (VCS) - git:} A brief discussion of
     \lq\lq git\rq\rq\ will be taken up in class to show how to work and
     collaborate with other students and your instructor. The \lq\lq git\rq\rq\ 
     platform allows you to work on a laptop at home and then \lq\lq push\rq\rq\ 
     your work and any modifications to Github where the instructor can get to
     the work.
       [> > go there](https://jvkoebbe.github.io/math4610/lectures/lecture_02/md/git_primer)
       | [(pdf)](https://jvkoebbe.github.io/math4610/lecture_02/pdf/git_primer.pdf)
  \item {\bf Code for Determining Machine Precision:} A simple algorithm for
     determining the precision for number representation will be covered. Two
     different versions of the code will be presented that determine single and
     double precision for your computer. Note that these two codes will be used
     to show students how to build a shared library.
       [> > go there](https://jvkoebbe.github.io/math4610/lecture_02/html/finite_precision)
       | [(pdf)](https://jvkoebbe.github.io/math4610/lecture_02/pdf/git_primer.pdf)

  \item {\bf Building a Shared Library for Reusing Code:} In important skill
     that computational mathematicians should learn is the ability to take
     relatively small codes and turning these into reusable object files and
     then collecting the object files into a shared library. 
       [> > go there](https://jvkoebbe.github.io/math4610/lecture_02/html/finite_precision)
       | [(pdf)](https://jvkoebbe.github.io/math4610/lecture_02/pdf/git_primer.pdf)
        \href{https://jvkoebbe.github.io/math4610/lectures/lecture_01/pdf/github_primer.pdf}

  \item {\bf Testing the Codes in a Shared Library:} We will go over how to
     write a code that uses the object routines in a shared library. Linking a
     main code to a library will be covered.
       [> > go there](https://jvkoebbe.github.io/math4610/lecture_02/html/finite_precision)
       | [(pdf)](https://jvkoebbe.github.io/math4610/lecture_02/pdf/git_primer.pdf)

  \item {\bf Floating Point Representation of Numbers on Computers:} The IEEE
     standard for number representation will be covered. Number precision for
     32-bit and 64-bit systems will be covered. 
