\documentclass[10pt,fleqn]{article}
\usepackage{hyperref}
\usepackage{graphicx}


\setlength{\topmargin}{-.75in}
\addtolength{\textheight}{2.00in}
\setlength{\oddsidemargin}{.00in}
\addtolength{\textwidth}{.75in}

\nofiles

\pagestyle{empty}

\setlength{\parindent}{0in}

\input{/cygdrive/m/tex/commands/commands}

\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vskip0.1in\hrule\vskip0.1in \noindent
{\bf Math 4610 Fundamentals of Computational Mathematics  - Topic 15.}
\vskip0.1in\hrule\vskip0.1in \noindent
In this section a brief introduction to Markdown will be presented. Since most
of the work we will do in the course will be appear on Github and Github likes
Markdown as a web page formatting tool it makes some sense to be familiar with
at least a few basics. Github has a number of guides that go through various
parts of Github like a handbook on git, using Github Pages, workflows and other
topics needed to master Github.

The location of the guides is given by
\vskip0.1in\hrule\vskip0.1in \noindent
\href{https://guides.github.com}{Github Guides}
\vskip0.1in\hrule\vskip0.1in \noindent
These guides provide lots of good examples for all parts of Github. There is
even a \lq\lq\ Hello World\rq\rq\ tutorial that will help you create and use
repositories.

Our goal in this section is to look through Markdown on Github. We will also
look into Github Pages to display your work in Markdown.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vskip0.1in\hrule\vskip0.1in \noindent
{\bf Mastering Markdown:}
\vskip0.1in\hrule\vskip0.1in \noindent
One can link to the Github Guide,
\href{https://guides.github.com/features/mastering-markdown/}{Mastering Github}
and look through the documentation. In the title bar at the given location there
is an estimate of how long it will take you to read through the tutorial on
markdown. We will spend just a bit more time on the topic. The good news is that
the guides will be there for your reference for a long time. So, if you forget
something, you can always pick up the details in the Mastering Markdown guide.

Note that we are going through this guide for two reasons. The first reason is
that Github uses Markdown and the guide will give us this start. The second
reason is that there is no advantage in reinventing the wheel in this case.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vskip0.1in\hrule\vskip0.1in \noindent
{\bf Drawbacks in Using Markdown:}
\vskip0.1in\hrule\vskip0.1in \noindent
Markdown is very simple to use, but does have a few drawbacks. First, the
extensive types of formatting using HTML and/or Cascading Style Sheets (CSS) are
not available. It is a quick and dirty way to create good looking web pages from
simple content. If the content is more complicated or needs some touch ups with
CSS it is more difficult to use.

Another advantage is that Markdown has easy formatting for different coding
languages, like C, Javascript, etc. However, if you are embedding mathematical
text, using Markdown will
not work. Javascript based extensions like MathJax do not play well with any
Markdown files. You can do some work to get \LaTeX to display mathematics. It
is just a lot easier to use HTML in this case.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vskip0.1in\hrule\vskip0.1in \noindent
{\bf Github Pages:}
\vskip0.1in\hrule\vskip0.1in \noindent
There is also a guide for GitHub Pages which will provide formatting through
Jekyll styles. You can have Github do all of this formatting work if you follow
the instructions in \href{https://guides.github.com/features/pages/}
{Getting Started with GitHub Pages} We will work through this guide to formalize
the ideas in using GitHub Pages.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}
