\documentclass[10pt,fleqn]{article}
\usepackage{hyperref}
\usepackage{graphicx}

\setlength{\topmargin}{-.75in}
\addtolength{\textheight}{2.00in}
\setlength{\oddsidemargin}{.00in}
\addtolength{\textwidth}{.75in}

\nofiles

\pagestyle{empty}

\setlength{\parindent}{0in}

\input{/cygdrive/m/tex/commands/commands}

\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vskip0.1in\hrule\vskip0.1in \noindent
{\bf{\Large Math 4610 Fundamentals of Numerical Analysis Tasksheet 2}}
\vskip0.1in\hrule\vskip0.1in \noindent
The problems for the Tasksheet 02 are included below. The deadline for turning
in your work on these problems will be posted on the repository. In addition,
you will turn in your work through the math4610 repository. A directory will be
constructed that will be used as a place to store your work.
\vskip0.1in\hrule\vskip0.1in \noindent
{\bf{\large Tasks}}
\vskip0.1in\hrule\vskip0.1in \noindent
\begin{enumerate}
\item {\bf Task 1:} After deciding which programming language you will use,
      write a program that provides a response to the "Hello World!" output.
      Provide details of the compilation and execution of the code you write.
      You can choose a suitable response - say something like..."it's only a
      bunny..."
\item {\bf Task 2:} Edit the main README.md file in your Math 4610 repository on
      Github using your favorite browser. In the README.md file, create an
      introduction that describes what the repository is being created for and
      put in a link to the table of contents for the homework problems and a
      link to the software manual you will create. Do this in Markdown.
\item {\bf Task 3:} Create the table of contents in a file with a name like
      \begin{verbatim}

        task_toc.md

      \end{verbatim}
      or something like this.
\item {\bf Task 4:} Write out the analysis for the centered difference
      approximation in a Taylor series expansions. Show that the approximation
      is second order.
\item {\bf Task 5:} Determine the order of accuracy of the central difference
      approximation of the second derivative. That is, analyze
      \[
        f'(x) \approx {{f(x+h) - 2 f(x) + f(x-h)}\over{h}}
      \]
      Write a code that approximates the second derivative of the function
      \[
        f(x) = cos(x)
      \] 
      at the point, \(x=2.0\).
\item {\bf Task 6:} Search the internet for information regarding finite
      difference approximations of derivatives of different orders. Give
      examples of these types of approximations. You should be able to find many
      such examples. Write a brief paragraph (3 or 4 sentences) that describe
      your findings. Include links to the sites you cite.
\end{enumerate}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}
