\documentclass[10pt,fleqn]{article}
\usepackage{hyperref}

\setlength{\topmargin}{-.75in}
\addtolength{\textheight}{2.00in}
\setlength{\oddsidemargin}{.00in}
\addtolength{\textwidth}{.75in}

\nofiles

\pagestyle{empty}

\setlength{\parindent}{0in}

\input{/cygdrive/m/tex/commands/commands}

\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\noindent
{\bf{\Large Introduction to the Course }}
\vskip0.1in\hrule\vskip0.1in
\noindent
{\bf Topic 1: Math 4610 Fundamentals of Numerical Analysis} 
\vskip0.1in\hrule\vskip0.1in
\noindent
This course presents fundamental content from numerical methods/analysis and
and introductory computational skills. Methods and some analysis of the methods
will be presented in the lectures. The content for each lecture in Math 4610
will be presented as a subset from a list of topics. The topics will be covered
in order as a continuous stream of material until the end of the semester. So,
we may cover one or two topics during a single lecture or we may take multiple
lectures to cover a single topic.

The course will not cover a long and thorough list of topics from computational
mathematics. Instead of covering a long list of numerical methods, we will weave
in computational skills like parallel programming, data visualization, and a few
topics that are not ordinarily covered in a numerical analysis or numerical
methods course.

The next step by going over a syllabus for the course.

\vskip0.1in\hrule\vskip0.1in \noindent
  \href{../../guided_notes/topics/toc/pdf/embed_toc.md}{} \|
  \href{./embed_xx.md}{Table of Contents} \|
  \href{./embed_xx.md}{Next}
\vskip0.1in\hrule\vskip0.1in \noindent

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}
